---
title: "Causal Effects"
author: "Dimitris Rizopoulos"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Causal Effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library("JMbayes2")
library("lattice")
```

# Causal Effects from Joint Models
We will illustrate the calculation of causal effects from joint models using the PBC dataset for the longitudinal outcome `serBilir` and the composite event transplantation or death. We start by fitting a joint model to the data. In the longitudinal submodel, we specify nonlinear subject-specific trajectories using natural cubic splines. In the fixed-effects part, we also include the treatment effect and its interaction with time. In the survival submodel, we only include the treatment effect.
```{r, "joint model PBC data"}
pbc2.id$status2 <- as.numeric(pbc2.id$status != "alive")
lmeFit <- lme(log(serBilir) ~ ns(year, 3, B = c(0, 14.4)) * drug, 
                   data = pbc2, random = ~ ns(year, 3, B = c(0, 14.4)) | id,
                   control = lmeControl(opt = "optim"))
CoxFit <- coxph(Surv(years, status2) ~ drug, data = pbc2.id)
jmFit <- jm(CoxFit, lmeFit, time_var = "year")
summary(jmFit)
```

The coefficient for `drugD-penicil` for the survival outcome in the output produced by the `summary()` method denotes the residual/direct effect of treatment on the risk of the composite event. It does not include the effect of treatment that follows via the serum bilirubin pathway.

We will illustrate the calculation of causal risk differences for the group of patients that have the same distribution of serum bilirubin values as Patient 2:
```{r, fig.align = "center", fig.width = 8.5, fig.height = 7.5}
xyplot(log(serBilir) ~ year, data = pbc2, subset = id == 2, type = "b",
       xlab = "Follow-up time (years)", ylab = "log{serum bilirubin (mg/dL)}",
       main = "Patient 2")
```

We calculate the risk difference for the composite event between the active treatment D-penicillamine and placebo at the horizon time `t_horiz = 6` using the longitudinal data up to year `t0 = 4`. To achieve this, we create a dataset with this patient's data. This patient received the active treatment D-penicillamine; hence, we also create a version of her data with the `drug` variable set to `placebo`:
```{r, "Patient 2 data"}
t0 <- 4
t_horiz <- 6
dataP2_Dpenici <- pbc2[pbc2$id == 2 & pbc2$year <= t0, ]
dataP2_Dpenici$years <- t0
dataP2_Dpenici$status2 <- 0

dataP2_placebo <- dataP2_Dpenici
dataP2_placebo$drug <- factor("placebo", levels = levels(pbc2$drug))
```

Note that in the `dataP2_placebo` dataset, we need to specify that `drug` is a factor with two levels. We also specify that the last time point we know the patient was still event-free was `t0`.

We estimate the cumulative risk for the composite event at `t_horiz` under the active treatment arm using the `predict()` method:
```{r, "Risk active treatment"}
Pr1 <- predict(jmFit, newdata = dataP2_Dpenici, process = "event", 
               times = t_horiz, return_mcmc = TRUE)
```
 
We have set the argument `return_mcmc` to `TRUE` to enable the calculation of a credible interval that accounts for the MCMC uncertainty. We produce the same estimate under the placebo arm:
```{r, "Risk placebo"}
Pr0 <- predict(jmFit, newdata = dataP2_placebo, process = "event", 
               times = t_horiz, return_mcmc = TRUE)
```

The estimated risk difference and its 95% credible interval are calculated by the corresponding elements of the `Pr1` and `Pr0` objects, i.e.,
```{r, "risk difference"}
# estimate 
Pr1$pred[2L] - Pr0$pred[2L]

# MCMC variability
quantile(Pr1$mcmc[2L, ] - Pr0$mcmc[2L, ], probs = c(0.025, 0.975))
```

## Time-varying treatments 
An extended example with time-varying treatments / intermediate events that showcases a calculation of the variance of the causal effects that includes the sampling variability is available [here](https://github.com/drizopoulos/JMbayes2/blob/master/Development/CI/causal_effects.R).
