---
title: "Multi-State Processes"
author: "Grigorios Papageorgiou"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Multi-State Processes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library('JMbayes2')
library('MASS')
```

# Joint Models with Multi-state Processes
## Introduction
A subject may often transition between multiple states, and we are interested in assessing the association of longitudinal marker(s) with each of these transitions. This vignette illustrates how to achieve this using **JMbayes2**. 

We will consider a simple case with one longitudinal outcome and a three-state (illness-death) model, but this application can be extended for the cases of multiple longitudinal markers and more than three states. 

## Data
First, we will simulate data from a joint model with a single linear mixed effects model and a multi-state process with three possible states. The multi-state process can be visualized as:
```{r ms_figure, echo = FALSE, warning = FALSE, message = FALSE}
library("ggplot2")

d <- data.frame("xmin" = c(0, 45, 22.5), "xmax" = c(15, 60, 37.5), 
                "ymin" = c(50, 50, 0), "ymax" = c(60, 60, 15))

dline <- data.frame("x1" = c(15), "y1" = c(55), "xend" = c(45), "yend" = c(55))

dcurve <- data.frame("x1" = c(7.5, 52.5), "y1" = c(50, 50), 
                     "xend" = c(22.4, 37.4), "yend" = c(7.5, 7.5), 
                     "curvature" = c(0, 0))

ggplot() + geom_rect(data = d, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), 
                     fill = "#ffffff", color = 'black', 
                     size = 1) + 
  geom_text(aes(x = 7.5, y = 55, label = "Healthy"), size = 6.5) +
  geom_text(aes(x = 52.5, y = 55, label = "Illness"), size = 6.5) +
  geom_text(aes(x = 30, y = 7, label = "Death"), size = 6.5) +
  geom_text(aes(x = 30, y = 52.5, label = 'h[1][2](t)'), size = 4, parse = TRUE) +
  geom_text(aes(x = 12.5, y = 30, label = "h[1][3](t)"), size = 4, parse = TRUE) +
  geom_text(aes(x = 47.5, y = 30, label = "h[2][3](t)"), size = 4, parse = TRUE) +
  geom_segment(data = dline, aes(x = x1, y = y1, xend = xend, yend = yend), size = 1, 
               arrow = arrow(length = unit(0.02, "npc"))) + 
  geom_curve(data = dcurve[1, ], aes(x = x1, y = y1, xend = xend, yend = yend), 
             size = 1, curvature = 0, 
             arrow = arrow(length = unit(0.02, "npc"))) +
  geom_curve(data = dcurve[2, ], aes(x = x1, y = y1, xend = xend, yend = yend), 
             size = 1, curvature = 0, 
             arrow = arrow(length = unit(0.02, "npc"))) +
  ylim(0, 60) + xlim(0, 60) + 
  theme_void()
```

where all subjects start from the state "Healthy" and then can transition to either state "Illness" and then state "Death" or directly to state "Death." In this case, states "Healthy" and "Illness" are *transient* states as the subject, when occupying these states, can still transition to other states, whereas "Death" is an absorbing state as when a subject reaches this state, then no further transitions can occur. This means that three transitions are possible: $1 \rightarrow 2$, $1 \rightarrow 3$ and $2 \rightarrow 3$ with transition intensities $h_{12}\left(t\right)$, $h_{13}\left(t\right)$ and $h_{23}\left(t\right)$ respectively. 

For our example, the default functional form is assumed, i.e., that the linear predictor $\eta(t)$ of the mixed model is associated with each transition intensity at time $t$. The following piece of code simulates the data:
```{r "simulate_Multi_State"}
# Set seed for reproducibility
set.seed(1710)

# Sample sizes and settings
N <- 1500
n_per <- 14

# Baseline covariate (binary)
X <- rbinom(N, 1, 0.5)

# Longitudinal parameters
beta <- c(5, -0.1)
sigma_e <- 1
D <- matrix(c(1.0, 0.2, 0.2, 0.3), 2, 2)
b <- MASS::mvrnorm(N, mu = c(0, 0), Sigma = D)

# Simulate longitudinal data
df_long <- do.call(rbind, lapply(1:N, function(i) {
    obs_times <- sort(c(0, runif(n_per - 1, 0, 18)))
    XX <- cbind(1, obs_times)
    eta <- XX %*% beta + XX %*% b[i, ]
    y <- rnorm(n_per, eta, sigma_e)
    data.frame(id = i, time = obs_times, y = y, X = X[i])
}))

# Weibull hazard parameters (shape, scale), association, covariate effects
weib_shape <- c(2.2, 1.8, 2.5)
weib_scale <- c(0.04, 0.03, 0.05)
alpha <- c(0.6, 0.4, 0.8) # different for each transition
gamma <- c(0.5, -0.3, 0.7) # different for each transition
Ctimes <- runif(N, 7, 10)

# Longitudinal trajectory function
yfun <- function(b_i, t) beta[1] + b_i[1] + (beta[2] + b_i[2]) * t

# Simulate event time via inverse transform
sim_time <- function(shape, scale, alpha, gamma, X, b_i, tmax = 20) {
    U <- runif(1)
    cumhaz <- function(t) {
        integrate(function(s) {
            h0 <- shape * scale * (scale * s)^(shape - 1)
            h0 * exp(alpha * yfun(b_i, s) + gamma * X)
        }, 0, t)$value
    }
    f <- function(t) cumhaz(t) + log(U)
    out <- try(uniroot(f, c(1e-5, tmax))$root, silent = TRUE)
    if (inherits(out, "try-error")) Inf else out
}

# Simulate multi-state transitions
df_surv <- do.call(rbind, lapply(1:N, function(i) {
    b_i <- b[i, ]; X_i <- X[i]; C <- Ctimes[i]
    T01 <- sim_time(weib_shape[1], weib_scale[1], alpha[1], gamma[1], X_i, b_i)
    T02 <- sim_time(weib_shape[2], weib_scale[2], alpha[2], gamma[2], X_i, b_i)
    
    if (T01 < T02 && T01 < C) {
        T12 <- sim_time(weib_shape[3], weib_scale[3], alpha[3], gamma[3], X_i, b_i)
        Tdeath <- T01 + T12
        data.frame(id = i, 
                   transition = c(1, 2, 3), 
                   Tstart = c(0, 0, T01), 
                   Tstop = c(T01, T01, min(Tdeath, C)), 
                   status = c(1, 0, as.integer(T12 < (C - T01))), 
                   X = X_i)
    } else if (T02 < C) {
        data.frame(id = i, transition = c(1, 2), Tstart = c(0, 0), Tstop = c(T02, T02), status = c(0, 1), X = X_i)
    } else {
        data.frame(id = i,
                   transition = c(1, 2),
                   Tstart = 0,
                   Tstop = C,
                   status = 0,
                   X = X_i)
    }
}))

# Filter longitudinal data to observed period
df_long2 <- merge(df_long, df_surv[, c("id", "Tstop")], by = c("id"))
df_long2 <- df_long2[!duplicated(df_long2, by = c('id', 'time')), ]
df_long2 <- df_long2[df_long2$time <= df_long2$Tstop, ]

```

The data for the multi-state process need to be in the appropriate long format:
```{r}
head(df_surv, n = 5L)
```

For example, subject 1 experienced the following transition: $1 \rightarrow 2$ and therefore is represented in 3 rows, one for each transition, because all of these transitions were plausible. On the other hand, subject 2 is only represented by two rows, only for transitions $1 \rightarrow 2$ and $1 \rightarrow 3$ since these are the only transitions possible from state 1. Since subject 2 never actually transitioned to state 2, transition $2 \rightarrow 3$ was never possible, and therefore, no row for this transition is in the dataset. It is also important to note that the time in the dataset follows the counting process formulation with intervals specified by `Tstart` and `Tstop` and that there is a variable (in this case `transition`) indicating which transition the row corresponds to.

## Fitting the model
When the data in the appropriate format are available, fitting the model is very straightforward. First we fit a linear mixed model using the `lme()` function from package **nlme**:
```{r, "mixed_model_fit"}
mixedmodel <- lme(y ~ time, random = ~ time | id, data = df_long2)
```

Then, we fit a multi-state model using function `coxph()` from package **survival**, making sure we use the counting process specification and that we add `strata(transition)` to stratify by the transition indicator variable in the dataset. Furthermore, we add an interaction between covariate `X` and each transition to allow the effect of this covariate to vary across transitions.
```{r, "multi_state_model_fit"}
msmodel <- coxph(Surv(Tstart, Tstop, status) ~ X:strata(transition), data = df_surv)
```

Finally, to fit the joint model, we simply run:
```{r, "jm_fit"}
jm_ms_model <- jm(msmodel, mixedmodel, time_var = "time", n_iter = 5000L,
                  functional_forms = ~ value(y):strata(transition))

summary(jm_ms_model)
```

which differs from a default call to `jm()` by the addition of the `functional_forms` argument specifying that we want an "interaction" between the marker's value and each transition, which translates into a separate association parameter for the longitudinal marker and each transition.
